\stktokens{}: Enforcing Well-bracketed Control Flow and Stack Encapsulation using Linear Capabilities

### 2018-07-05 fully-abstract-comp/popl2019/paper.tex
minor changes to trg machine section.

At this time, the executable configuration is our only way to specify the state of the capability machine.
We would, however, much prefer a notion of a program that only describe the essential parts of an entity on the machine and only later instrument the program with whatever necessary to make it into an executable configuration.
In later sections, we describe different entities on the capability machine that interact with each other.
In order for them to interact, we need to be able to link them together and to this end, we introduce a notion of a component.
A component is basically a program with open ends in the form of imports.
A component also have exports allowing other programs to link with them.
We define the notion of a component in Figure~\ref{fig:target-component-and-linking}.
% Structure
A base component $\comp_0$ consists of the following: a code memory segment, a data memory segment, a list of imported symbols, a list of exported symbols, two lists specifying the available seals seals\footnote{We will return to the seals in later sections.}, and a set of all the linear addresses.
The import list specifies where in memory imports should be placed, and imports are matched to exports via their symbols.
The exports are words associated with a symbol.
%base = library
A base component can be seen as library implementations because it does not specify a main entry point of execution.
Components are either a library component or a base component with a main in the form of a pair of sealed capabilities which can be seen as a program that still needs to be linked with libraries.
% Linking .
Components are combined into new components by linking them together.
Libraries can be linked together and incomplete programs can be linked with libraries, but two incomplete programs cannot be linked with each other.
Two components can be linked when their memories, seals, and linear addresses disjointed.
They are combined by taking the union of each of their constituents.
For every import that is satisfied by an export of the other component, in the sense that they have the same symbol, the data memory is updated to have the exported word on the imported address.
The satisfied imports are removed from the import list in the resulting linked component and the exports are updated to be the exports of the two components.
Linking is defined in Figure~\ref{fig:target-component-and-linking}.

% Programs
We can now define the notion of a program as well as a context.
\begin{definition}[Programs and Contexts]
  We define a \emph{program} to be a component $(\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d})$ with an empty import list.

  A \emph{context} for a component $\var{comp}$ is another component $\var{comp}'$ such that $\var{comp} \bowtie \var{comp}'$ is a program.
\end{definition}
% Getting an executable from a component - postpone this to later, but mention:
% write xor exec
In Section~\ref{sec:form-secur-with}, we will get back to what a program is instrumented with in order to create an executable configuration. 
We will, however, at this point mention that all programs will be instrumented with the Write-XOR-Execute,\todo{Do we have a reference for this?} i.e.\ there will only be read-execute capabilities for the code and read-write capabilities for the data.
%Further, we will only consider a specific class of well-formed components, but we will get back to this.

\begin{figure}[htb]
  \centering
  \[
    \begin{array}{rl c rl c rl}
    s &\in \Symbol & \phantom{make} &
    \mathrm{import} &\mathrel{::=} a \mapsfrom s & \phantom{make} &
    \mathrm{export} &\mathrel{::=} s \mapsto w\\
    \var{comp}_0 & \multicolumn{7}{l}{\mathrel{::=} (\mscode,\msdata,\overline{\mathrm{import}},\overline{\mathrm{export}},\sigrets,\sigcloss,A_\linear)}\\
    \var{comp} & \multicolumn{7}{l}{\mathrel{::=} \var{comp}_0 \mid  (\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d})}
  \end{array}
\]
\begin{mathpar}
  \inference{
    \var{comp}_1 = (\mscode[1], \msdata[1], \overline{\var{import}_1}, \overline{\var{export}_1}, \sigrets[1], \sigcloss[1],A_{\linear,1})\\
    \var{comp}_2 = (\mscode[2], \msdata[2], \overline{\var{import}_2}, \overline{\var{export}_2}, \sigrets[2], \sigcloss[2],A_{\linear,2})\\
    \var{comp}_3 = (\mscode[3], \msdata[3], \overline{\var{import}_3}, \overline{\var{export}_3}, \sigrets[3], \sigcloss[3],A_{\linear,3})\\
    \mscode[3] = \mscode[1] \uplus \mscode[2] \\
    \msdata[3] = (\msdata[1] \uplus \msdata[2])[a \mapsto w \mid (a \mapsfrom s) \in (\overline{\var{import}_1} \cup \overline{\var{import}_2}), (s \mapsto w) \in \overline{\var{export}}_3] \\
    \overline{\var{export}_3} = \overline{\var{export}_1} \cup \overline{\var{export}_2}&
    \overline{\var{import}_3} = \{ a \mapsfrom s \in (\overline{\var{import}_1} \cup \overline{\var{import}_2}) \mid s \mapsto \_ \not\in \overline{\var{export}_3} \}\\
    \sigrets[3] = \sigrets[1] \uplus \sigrets[2] &
    \sigcloss[3] = \sigcloss[1] \uplus \sigcloss[2] &
    A_{\linear,3} = A_{\linear,1} \uplus A_{\linear,2}\\
    \dom(\mscode[3]) \mathrel{\#} \dom(\msdata[3]) & \sigrets[3] \mathrel{\#} \sigcloss[3]
  } {
    \var{comp}_3 = \var{comp}_1 \bowtie \var{comp}_2
  }
  \and
  \inference{
    \var{comp}_0'' = \var{comp}_0 \bowtie \var{comp}_0'
  }{
    (\var{comp}_0'',c_{\mathrm{main},c}, c_{\mathrm{main},d}) = \var{comp}_0 \bowtie (\var{comp}_0',c_{\mathrm{main},c}, c_{\mathrm{main},d}) = (\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d}) \bowtie \var{comp}_0'
  }
\end{mathpar}
% \begin{definition}[Initial execution configuration]
%   \begin{mathpar}
%     \inference{
%       c_{\mathrm{main},c} = \sealed{\sigma_1, c_{\mathrm{main},c}'} &
%       c_{\mathrm{main},d} = \sealed{\sigma_2, c_{\mathrm{main},d}'} &
%       \sigma_1 = \sigma_2\\
%       \nonExec{c_{\mathrm{main},d}'}&
%       \reg(\pcreg) = c_{\mathrm{main},c}' & 
%       \reg(\rdata) = c_{\mathrm{main},d}' \\
%       \src{\reg(\rstk) = \stkptr{\rw,\baddr_\stk,\eaddr_\stk,\eaddr_\stk}} & 
%       \trg{\reg(\rstk) = ((\rw,\linear),\baddr_\stk,\eaddr_\stk,\eaddr_\stk)} \\
%       \reg(\RegName \setminus \{\pcreg,\rdata,\rstk\}) = 0\\
%       \range{\ms_\stk} = \{0\}&
%       \mem = \mscode \uplus \msdata \trg{\;\uplus\; \ms_\stk} \\
%       [\baddr_\stk,\eaddr_\stk] = \dom(\ms_\stk ) \mathrel{\#} (\dom(\mscode) \cup \dom(\msdata)) &
%       \overline{\var{import}} = \emptyset
%     }{
%       ((\mscode, \msdata, \overline{\var{import}}, \overline{\var{export}}, \sigrets, \sigcloss,A_\linear),  c_{\mathrm{main},c}, c_{\mathrm{main},d}) \rightsquigarrow (\mem, \reg\src{, \emptyset, \ms_\stk})
%     }
%   \end{mathpar}
% \end{definition}
% \begin{definition}[Plugging a program into a context]
%   When $\var{comp'}$ is a context for component $\var{comp}$ and $\var{comp}' \bowtie \var{comp} \rightsquigarrow \Phi$, 
%   then we write $\plug{\var{comp'}}{\var{comp}}$ for the execution configuration $\Phi$.
% \end{definition}
  \caption{Components and linking of components}
  \label{fig:target-component-and-linking}
\end{figure}

Some simplifications has been made in this presentation of \trgcm{} in this section. We refer to the accompanying technical report\citep{technical_report} for all details.

>>>>

At this time, the executable configuration is our only way to specify the state of the capability machine.
We would, however, much prefer a notion of a program that only describe the essential parts of an entity on the machine and only later instrument the program with whatever necessary to make it into an executable configuration.
In later sections, we describe different entities on the capability machine that interact with each other.
In order for them to interact, we need to be able to link them together and to this end, we introduce a notion of a component.
A component is basically a program with open ends in the form of imports.
A component also have exports allowing other programs to link with them.
We define the notion of a component in Figure~\ref{fig:target-component-and-linking}.
% Structure
A base component $\comp_0$ consists of the following: a code memory segment, a data memory segment, a list of imported symbols, a list of exported symbols, two lists specifying the available seals seals\footnote{We will return to the seals in later sections.}, and a set of all the linear addresses.
The import list specifies where in memory imports should be placed, and imports are matched to exports via their symbols.
The exports are words associated with a symbol.
%base = library
A base component can be seen as library implementations because it does not specify a main entry point of execution.
Components are either a library component or a base component with a main in the form of a pair of sealed capabilities which can be seen as a program that still needs to be linked with libraries.
% Linking .
Components are combined into new components by linking them together.
Libraries can be linked together and incomplete programs can be linked with libraries, but two incomplete programs cannot be linked with each other.
Two components can be linked when their memories, seals, and linear addresses disjointed.
They are combined by taking the union of each of their constituents.
For every import that is satisfied by an export of the other component, in the sense that they have the same symbol, the data memory is updated to have the exported word on the imported address.
The satisfied imports are removed from the import list in the resulting linked component and the exports are updated to be the exports of the two components.
In Figure~\ref{fig:target-component-and-linking}, $\bowtie$ defines linking.

% Programs
We can now define the notion of a program as well as a context.
\begin{definition}[Programs and Contexts]
  \label{def:program-and-context}
  We define a \emph{program} to be a component $(\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d})$ with an empty import list.

  A \emph{context} for a component $\var{comp}$ is another component $\var{comp}'$ such that $\var{comp} \bowtie \var{comp}'$ is a program.
\end{definition}
% Getting an executable from a component - postpone this to later, but mention:
% write xor exec
In Section~\ref{sec:form-secur-with}, we will get back to what a program is instrumented with in order to create an executable configuration. \todo[inline]{Make sure we got back to this!}
We will, however, at this point mention that all programs will be instrumented with the Write-XOR-Execute,\todo[inline]{Do we have a reference for this?} i.e.\ there will only be read-execute capabilities for the code and read-write capabilities for the data.
%Further, we will only consider a specific class of well-formed components, but we will get back to this.

Some simplifications has been made in this presentation of \trgcm{} in this section. We refer to the accompanying technical report\citep{technical_report} for all details.
\begin{figure}[htb]
  \centering
  \[
    \begin{array}{rl c rl c rl}
    s &\in \Symbol & \phantom{make} &
    \mathrm{import} &\mathrel{::=} a \mapsfrom s & \phantom{make} &
    \mathrm{export} &\mathrel{::=} s \mapsto w\\
    \var{comp}_0 & \multicolumn{7}{l}{\mathrel{::=} (\mscode,\msdata,\overline{\mathrm{import}},\overline{\mathrm{export}},\sigrets,\sigcloss,A_\linear)}\\
    \var{comp} & \multicolumn{7}{l}{\mathrel{::=} \var{comp}_0 \mid  (\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d})}
  \end{array}
\]
\begin{mathpar}
  \inference{
    \var{comp}_1 = (\mscode[1], \msdata[1], \overline{\var{import}_1}, \overline{\var{export}_1}, \sigrets[1], \sigcloss[1],A_{\linear,1})\\
    \var{comp}_2 = (\mscode[2], \msdata[2], \overline{\var{import}_2}, \overline{\var{export}_2}, \sigrets[2], \sigcloss[2],A_{\linear,2})\\
    \var{comp}_3 = (\mscode[3], \msdata[3], \overline{\var{import}_3}, \overline{\var{export}_3}, \sigrets[3], \sigcloss[3],A_{\linear,3})\\
    \mscode[3] = \mscode[1] \uplus \mscode[2] \\
    \msdata[3] = (\msdata[1] \uplus \msdata[2])[a \mapsto w \mid (a \mapsfrom s) \in (\overline{\var{import}_1} \cup \overline{\var{import}_2}), (s \mapsto w) \in \overline{\var{export}}_3] \\
    \overline{\var{export}_3} = \overline{\var{export}_1} \cup \overline{\var{export}_2}&
    \overline{\var{import}_3} = \{ a \mapsfrom s \in (\overline{\var{import}_1} \cup \overline{\var{import}_2}) \mid s \mapsto \_ \not\in \overline{\var{export}_3} \}\\
    \sigrets[3] = \sigrets[1] \uplus \sigrets[2] &
    \sigcloss[3] = \sigcloss[1] \uplus \sigcloss[2] &
    A_{\linear,3} = A_{\linear,1} \uplus A_{\linear,2}\\
    \dom(\mscode[3]) \mathrel{\#} \dom(\msdata[3]) & \sigrets[3] \mathrel{\#} \sigcloss[3]
  } {
    \var{comp}_3 = \var{comp}_1 \bowtie \var{comp}_2
  }
  \and
  \inference{
    \var{comp}_0'' = \var{comp}_0 \bowtie \var{comp}_0'
  }{
    (\var{comp}_0'',c_{\mathrm{main},c}, c_{\mathrm{main},d}) = \var{comp}_0 \bowtie (\var{comp}_0',c_{\mathrm{main},c}, c_{\mathrm{main},d}) = (\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d}) \bowtie \var{comp}_0'
  }
\end{mathpar}
% \begin{definition}[Initial execution configuration]
%   \begin{mathpar}
%     \inference{
%       c_{\mathrm{main},c} = \sealed{\sigma_1, c_{\mathrm{main},c}'} &
%       c_{\mathrm{main},d} = \sealed{\sigma_2, c_{\mathrm{main},d}'} &
%       \sigma_1 = \sigma_2\\
%       \nonExec{c_{\mathrm{main},d}'}&
%       \reg(\pcreg) = c_{\mathrm{main},c}' & 
%       \reg(\rdata) = c_{\mathrm{main},d}' \\
%       \src{\reg(\rstk) = \stkptr{\rw,\baddr_\stk,\eaddr_\stk,\eaddr_\stk}} & 
%       \trg{\reg(\rstk) = ((\rw,\linear),\baddr_\stk,\eaddr_\stk,\eaddr_\stk)} \\
%       \reg(\RegName \setminus \{\pcreg,\rdata,\rstk\}) = 0\\
%       \range{\ms_\stk} = \{0\}&
%       \mem = \mscode \uplus \msdata \trg{\;\uplus\; \ms_\stk} \\
%       [\baddr_\stk,\eaddr_\stk] = \dom(\ms_\stk ) \mathrel{\#} (\dom(\mscode) \cup \dom(\msdata)) &
%       \overline{\var{import}} = \emptyset
%     }{
%       ((\mscode, \msdata, \overline{\var{import}}, \overline{\var{export}}, \sigrets, \sigcloss,A_\linear),  c_{\mathrm{main},c}, c_{\mathrm{main},d}) \rightsquigarrow (\mem, \reg\src{, \emptyset, \ms_\stk})
%     }
%   \end{mathpar}
% \end{definition}
% \begin{definition}[Plugging a program into a context]
%   When $\var{comp'}$ is a context for component $\var{comp}$ and $\var{comp}' \bowtie \var{comp} \rightsquigarrow \Phi$, 
%   then we write $\plug{\var{comp'}}{\var{comp}}$ for the execution configuration $\Phi$.
% \end{definition}
  \caption{Components and linking of components}
  \label{fig:target-component-and-linking}
\end{figure}
